Qualitative Assessesment of Graph Neural Network Potentials for Phonon Properties of Shape Memory Alloys using Graph Neural Network Potentials

### 2024-01-30 src/tex/ms.tex
Plot density. MS edits.

\subsection{Workflow}
\label{sec:workflow}
To promote reproducible research and align with findable, accessible,
interoper-able and reusable (FAIR) principles\cite{Walsh2024}, we
utilize \showyourwork python workflow\cite{Luger2021} and several open
source python packages. Instructions for reproducing calculated data,
figures, tables, and manuscript are provided in
appendix~\ref{sec:appx_reproduce} and corresponding github
repo\footnote{\url{https://github.com/stefanbringuier/SMA_Phonons_GNNIP}}.\par

\subsection{Calculation Details}
\label{sec:calc_details}
The ground-state structures for the NiTi phases \textit{B2},
\textit{B19}, \textit{B19'}, and \textit{BCO} are generated using the
atomic simulation environment (ASE) package \cite{Larsen2017}. This
requires that the unit cell parameters, basis atoms, and spacegroup
number be provided. The initial unit cell parameters and basis
corrdinates for these structures have been taken from the materials
project \cite{Jain2013} and several references
\cite{Haskins2016,Kadkhodaei2018}. The initial structures for NiTi are
shown in fig~\ref{fig:niti_structures} and are adapted for PtTi as
well given same spacegroup symmetries. The energy and forces on the
atoms are determined using the ASE calculator interfaces for EAM,
MEAM, M3GNet, CHGNet, and
MACE.\cite{Mutter2010,Zhong2011,Ko2015,Kim2017,Chen2022,Deng2023,Batatia2022}
The EAM and MEAM potential interfaces are achieved through the use of
LAMMPS python library \cite{Thompson2022} and ASE. The ALIGNN ASE
calculator interface was also implemented, however, we found that the
predictions for NiTi and PtTi to be unreliable are not included in
this study.\par

For each combination of potential/model, chemical system, and phase we
perform calcualtions for the structure minimization, equation of state
(EOS), phonons, and elastic constants. To determine the
finite-temperature phase transitions we utlize an isothermal-isobaric
ensemble via molecular dynamics simulations. The thermostating and
barostating is done with Nose-Hoover and Parrinello-Rahman dynamics
and the tempertaure is adjusted in a quasi-static manner which entails
equlibrating the simulation for 10 ns while decrementing the
temperature by 10 Kelvin; the pressure is kept constant at 1 bar. We
note that in order to achieve the different computing environments
required by the GNN potentials we leverage the ability to specify the
python computing environments using
\href{https://github.com/showyourwork/showyourwork}{\showyourwork}
package.\par

\begin{figure}[ht!]
  \script{VisualizedStructures.py}
  \begin{centering}
    \includegraphics[width=\linewidth]{figures/NiTi_VisualizedStructures.png}
    \caption{ Initial \ce{NiTi} structures (i.e., unrelaxed) used for
      (a) \textit{B2}(221), (b) \textit{B19}(11), (c)
      \textit{B19'}(51), and (d) \textit{BCO}(61) with spacegroup
      numbers are given in the parentheses. The green atoms are
      \ce{Ni} and \ce{Ti} as gray and unit cells are highlighted with
      parametes shown on axis. For all unit cells the $\alpha$ and
      $\gamma$ angles are 90.0$^\circ$. For PtTi, (a-b) structures are
      used with different unit cell parameters and basis coordinates
      and then optimized accordingly.  }
    \label{fig:niti_structures}
  \end{centering}
\end{figure}

Prior to any property calculations or dynamics the structures are
optimized using the FIRE\cite{Bitzek2006} local optimizer with the
forces converged to $2\cdot 10^{-3}$eV/\AA. We observed that other
local optimization routines find the same ground-state structures. To
ensure that the unit cell parameters were adjusted along with the
atomic coordinates and that the crystal symmetry is not broken, we
utlize a constraint and filter during the optimization process. \par

In fig.~\ref{fig:opt_density} the density of the structures for
different potentials is shown. We observe that in general the
densities and lattice parameters are within agreement with previously
reported values. The optimized lattice parameters as well as basis
atoms for all structures and models are provided in appendix
\ref{sec:opt_structures}. \par

\begin{figure}[ht!]
  \label{fig:opt_density}
  \centering
  \caption{}
\end{figure}

@@

\subsubsection{Phonon Bandstructures}

@@

To better understand dynamic instability in the various structures,
the same isotropic strain in eq.~\ref{eq:isotropic_strain} is used and
the repsective phonon bandstructures are calculated. These are then
plotted to observed the onset or change in stability or instability of
phonon different branches. \par

@@

Similar to the phonon calculations the elastic constants within the
harmonic approximation the elastic constants can be determined from
linear-response of strain-stress calculations. Therefore, the
calculation requires the use of crystal symmetry along with atomic
displacements to determine respective elastic constant componenets. We
use the \textit{Elastic}\cite{Jochym2018} python package which
provides routines for generating the proper atomic displacement matrix
given specific crystal spacegroup. The elastic constants are
determined by solving a set of linear equations using singular value
decomposition and given calculated strains and stress. The
displacements occur with a min and max of 0.5\% and a total of 10
points are used. \par

@@

For application of EAM to shape memory alloys, and in particular NiTi,
\citet{Mutter2010} initially devised an Finnis-Sinclair style EAM
potential to capture the structural phase transitions between
\textit{B19'} and \textit{B2} phases. It was latter shown by
\cite{Zhong2011} that some adjustments to the EAM parameters was
needed to properly capture the finite-temperature displacement
dynamics for the martensitic transition \cite{Gur2017}. More recently
\citet{Ko2015} developed a second nearest neighbor(2NN) MEAM
parameterization of NiTi that demonstrated very good predictive
performance and improvements over ealier potentials.

@@

The use of ML potentials for SMA has been demonstrated by
\citet{Tang2022}. In their work, they utilized the
DeePMD-Kit\cite{Wang2018} neural network framework to parameterize a
descriptor matrix from atomic configurations that preserves
permutations and symmetries. The descriptors are then which feed into
a multilayer perceptron to calculate energies and forces. The neural
network interatomic potential demonstrated good accuracy of
ground-state properties with respect to DFT results and showed
agreement with experiments regarding temperature, stress, and defect
induced martensitic transformation. \paragraphmark

In the work of \citet{Tang2022} ground-state predictions for
\textit{B2}, \textit{B19'}, \textit{B19}, and
\textit{BCO}\footnote{Referred to as \textit{B33} prototype in
  \citet{Tang2022}} were provided and showed good agreement. However,
we note that deviation from Ni-Ti 1:1 stoichiometry or addition of a
third species was not investigated nor the inclusion of additional
alloying species. \par

The pre-trained GNN interatomic potentials explored in this work have
been parameterized over a large fraction of the interatomic potential
($>$ 80 elements) and therefore are not restricted to specific
chemistries or stoichiometries. They training data also includes
various phases for a given chemical system. This permits great
flexibility. We do not that fine-tuning, the process of adjusting GNN
weights and biases using additional training data, can be performed
with the GNN studied in this work. However, this is out of the scope
for this manuscript, given resources and time needed to generate such
training data for SMAs. \par

@@

One way to understand the phase stability for a given composition is
to compare the cohesive energy, which is defined as,

\begin{equation}
  \label{eq:cohesive_energy}
  E_{c} = \frac{E_{\text{bulk}}-\sum_i^N n_i E^_{\text{atom_i}}}{N}
\end{equation}

With $E_{\text{bulk}}$ the total energy of the system,
$\sum_i^N n_i E_{\text{atom_i}}$ is the sum over isolated atom
energies for $n_i$ species, and $N$ the number of atoms in bulk
reference. Classical interatomic potentials are offset with respect to
the isolated atom energies, so these terms are zero and neglected,
giving $E_{c} = \frac{E_{\text{bulk}}}{N}$. In fig.~\ref{fig:ecoh} the
bar chart for different phases and interatomic potentials models are
show. We compare these to dervided experimental values for
\textit{B2}\cite{Vandermause2024} and the value from The Materials
Project (MP) entry \texttt{MP-571} which corresponds to the
\textit{B2} phase \cite{MP--571}. In the figure we see that the EAM
and MEAM potentials show significantly lower cohesive energy compared
to the GNN potentials and are in better agreement with the
experimental value. Previous studies using the EAM and MEAM potentials
have been shown to reproduce previous experimental and DFT data
\cite{Haskins2016,Ko2015} such as cohesive energy. The GNNs predict a
more strongly bound and stable \textit{B2} structure but all are
consistent with the entry for The MP and calculation details can be
found in ref.~\cite{MP--571}. This consistency is resonable
considering all pretrained GNN models used in this work have been
developed with MP data. \par

@@

\begin{figure}[!htp]
  \script{PlotPhonons.py}
  \begin{centering}
    \includegraphics[width=\linewidth]{figures/NiTi_B2_ModelsPhononBandstructures.png}
    \caption{ Ground-state, no isotropic strain, Phonon bandstructure
      of B2 phase for all models. The Mutter and Zhong EAM potentials
      differ in that Mutter predicts no dynamic instability of the B2
      phase at 0K.  }
    \label{fig:allmodels_b2}
  \end{centering}
\end{figure}

\begin{figure}[!htp]
  \script{PlotPhonons.py}
  \begin{centering}
    \includegraphics[width=\linewidth]{figures/NiTi_B19_ModelsPhononBandstructures.png}
    \caption{ Ground-state phonon bandstructure of B19 phase for all
      models.  }
    \label{fig:allmodels_B19P}
  \end{centering}
\end{figure}

\begin{figure}[!htp]
  \script{PlotPhonons.py}
  \begin{centering}
    \includegraphics[width=\linewidth]{figures/NiTi_B19P_ModelsPhononBandstructures.png}
    \caption{ Ground-state phonon bandstructure of
      \textit{B19}$^\prime$ phase for all models.  }
    \label{fig:allmodels_B19P}
  \end{centering}
\end{figure}

\begin{figure}[!htp]
  \script{PlotPhonons.py}
  \begin{centering}
    \includegraphics[width=\linewidth]{figures/NiTi_BCO_ModelsPhononBandstructures.png}
    \caption{ Ground-state phonon bandstructure of BCO phase for all
      models.  }
    \label{fig:allmodels_B19P}
  \end{centering}
\end{figure}

@@

To better coberate the results in fig.~\ref{fig:modegruneisen}, the
phonon bandstructures as a result of isoptroic strain applied in
compression and tension are show in fig

\begin{figure}[!htp]
  \script{PlotPhonons.py}
  \begin{centering}
    \includegraphics[width=0.5\linewidth]{figures/NiTi_Mutter_B2_StrainsPhononBandstructures.png}
    \vspace{1mm}
    \includegraphics[width=0.5\linewidth]{figures/NiTi_Zhong_B2_StrainsPhononBandstructures.png}
    \vspace{1mm}
    \includegraphics[width=0.5\linewidth]{figures/NiTi_Ko_B2_StrainsPhononBandstructures.png}
    \caption{ Phonon dispersion of B2 structure for the (top) EAM
      Mutter, (middle) EAM Zhong, and (bottom) 2NN-MEAM Ko potentials
      at different isotropic strains. The optical branches have been
      made more transparent to emphasis the behavior in the TA$_1$ and
      TA$_2$ branches. The color indicates direction from compressive
      to tensile strain. Note that the strain percent shown is with
      respect to isotropic changes (i.e. volume).  }
    \label{fig:mutter_zhong_ko_phonon_b2}
  \end{centering}
\end{figure}

@@

\begin{figure}[!htp]
  \script{PlotPhonons.py}
  \begin{centering}
    \includegraphics[width=0.5\linewidth]{figures/NiTi_M3GNet_B2_StrainsPhononBandstructures.png}
    \vspace{1mm}
    \includegraphics[width=0.5\linewidth]{figures/NiTi_CHGNet_B2_StrainsPhononBandstructures.png}
    \vspace{1mm}
    \includegraphics[width=0.5\linewidth]{figures/NiTi_MACE_B2_StrainsPhononBandstructures.png}
    \caption{ Phonon dispersion of B2 structure for the (top) M3GNet,
      (middle) CHGNet, and (bottom) MACE potentials for different
      isotropic strains. Similar transcparency is used as done in
      \ref{fig:mutter_zhong_phonon_b2}.  }
    \label{fig:gnn_phonon_b2}
  \end{centering}
\end{figure}

>>>>

\subsection{Workflow}
\label{sec:workflow}
To promote reproducible research and align with findable, accessible,
interoper-able and reusable (FAIR) principles\cite{Walsh2024}, we
utilize \showyourwork python workflow\cite{Luger2021} and several open
source python packages. Instructions for reproducing calculated data,
figures, tables, and manuscript are provided in
appendix~\ref{sec:appx_reproduce} and corresponding github
repo\footnote{\url{https://github.com/stefanbringuier/SMA_Phonons_GNNIP}}. We
note that in order to achieve the different computing environments
required by the interatomic potentials we leverage the ability to specify the
python computing environments using \href{https://github.com/showyourwork/showyourwork}{\showyourwork}
package.\par

\subsection{Calculation Details}
\label{sec:calc_details}
The ground-state structures for the NiTi phases \textit{B2},
\textit{B19}, \textit{B19'}, and \textit{BCO} are generated using the
atomic simulation environment (ASE) package \cite{Larsen2017}. This
requires that the unit cell parameters, basis atoms, and spacegroup
number be provided. The initial unit cell parameters and basis
corrdinates for these structures have been taken from the materials
project \cite{Jain2013} and several references
\cite{Haskins2016,Kadkhodaei2018}. The initial structures for NiTi are
shown in fig~\ref{fig:niti_structures} and are adapted for PtTi as
well given same space group symmetries. The energy and forces on the
atoms are determined using the ASE calculator interfaces for EAM,
MEAM, M3GNet, CHGNet, and MACE.\cite{Mutter2010,Zhong2011,Ko2015,Kim2017,Chen2022,Deng2023,Batatia2022}
The EAM and MEAM potential interfaces are achieved through the use of
LAMMPS python library \cite{Thompson2022} and ASE. We have also implemented calculators for potentials by \cite{Kavousi2019}, \cite{Wang2018}, and \cite{Choudhary2021} but are not shown in the main manuscript of this work.\par

For each combination of potential/model, chemical system, and phase we
perform calculations for the structure minimization, equation of state
(EOS), phonons, and elastic constants. To determine the
finite-temperature phase transitions we utilize an isothermal-isobaric
ensemble via molecular dynamics simulations. The thermostating and
barostating is done with Nose-Hoover and Parrinello-Rahman dynamics
and the ramping down or up of temperature is adjusted in a dynamical manner
which entails adjusting the target thermostat temperature every few timesteps
based on the ramp-rate. The pressure is kept constant at 1 bar.

\begin{figure}[ht!]
	\script{VisualizedStructures.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_VisualizedStructures.png}
		\caption{ Initial \ce{NiTi} structures (i.e., unrelaxed) used for
			(a) \textit{B2}(221), (b) \textit{B19}(11), (c)
			\textit{B19'}(51), and (d) \textit{BCO}(61) with spacegroup
			numbers are given in the parentheses. The green atoms are
			\ce{Ni} and \ce{Ti} as gray and unit cells are highlighted with
			parameters shown on axis. For all unit cells the $\alpha$ and
			$\gamma$ angles are 90.0$^\circ$. For PtTi, (a-b) structures are
			used with different unit cell parameters and basis coordinates
			and then optimized accordingly.  }
		\label{fig:niti_structures}
	\end{centering}
\end{figure}

Prior to any property calculations or dynamics the structures are
optimized using the FIRE\cite{Bitzek2006} local optimizer with the
forces converged to $2\cdot 10^{-3}$eV/\AA\@. We observed that other
local optimization routines find the same ground-state structures. To
ensure that the unit cell parameters were adjusted along with the
atomic coordinates and that the crystal symmetry is not broken, we
utlize a constraint and filter during the optimization process. \par

@@

\subsubsection{Phonon band structure}

@@

To better understand dynamic instability in the various structures,
the same isotropic strain in eq.~\ref{eq:isotropic_strain} is used and
the respective phonon band structure are calculated. These are then
plotted to observed the onset or change in stability or instability of
phonon different branches. \par

@@

Similar to the phonon calculations the elastic constants within the
harmonic approximation the elastic constants can be determined from
linear-response of strain-stress calculations. Therefore, the
calculation requires the use of crystal symmetry along with atomic
displacements to determine respective elastic constant components. We
use the \textit{Elastic}\cite{Jochym2018} python package which
provides routines for generating the proper atomic displacement matrix
given specific crystal space group. The elastic constants are
determined by solving a set of linear equations using singular value
decomposition and given calculated strains and stress. The
displacements occur with a min and max of 0.5\% and a total of 10
points are used. \par

@@

For application of EAM to shape memory alloys, and in particular NiTi,
\citet{Mutter2010} initially devised an Finnis-Sinclair style EAM
potential to capture the structural phase transitions between
\textit{B19'} and \textit{B2} phases. It was latter shown by
\cite{Zhong2011} that some adjustments to the EAM parameters were
needed to properly capture the finite-temperature displacement
dynamics for the martensitic transition \cite{Gur2017}. More recently
\citet{Ko2015} developed a second nearest neighbor(2NN) MEAM
parameterization of NiTi that demonstrated very good predictive
performance and improvements over earlier potentials.

@@

The use of ML potentials for SMA has been demonstrated by
\citet{Tang2022}. In their work, they utilized the
DeePMD-Kit\cite{Wang2018} neural network framework to parameterize a
descriptor matrix from atomic configurations that preserves
permutations and symmetries. The descriptors are then feed into
a multilayer perceptron to calculate energies and forces. The neural
network interatomic potential demonstrated good accuracy of
ground-state properties with respect to DFT results and showed
agreement with experiments regarding temperature, stress, and defect
induced martensitic transformation. \paragraphmark

The ground-state predictions for
\textit{B2}, \textit{B19'}, \textit{B19}, and
\textit{BCO}\footnote{Referred to as \textit{B33} prototype in
	\citet{Tang2022}} in \citet{Tang2022} showed good agreement. However,
we note that deviation from Ni-Ti 1:1 stoichiometry or addition of a
third species was not investigated nor the inclusion of additional
alloying species was explored. \par

The pre-trained GNN interatomic potentials explored in this work have
been parameterized over a large fraction of the interatomic potential
($>$ 80 elements) and therefore are not restricted to specific
chemistries or stoichiometries. The training data also includes
various phases for a given chemical system which permits great
flexibility, although broad transferability still requires verification.
This allows for investigation of both non-stoichiometric and multi-species
SMA which previous physics and ML potentials were not developed for. We do
note that fine-tuning, the process of adjusting GNN weights and biases
using additional training data, can be performed
with the GNNs studied in this work. However, this is out of the scope
for this manuscript, given resources and time needed to generate such
training data for SMAs. \par

@@

In fig.~\ref{fig:opt_density} the density of the structures for
different potentials is shown. We observe that in general the
densities and lattice parameters are within agreement with previously
reported values. The optimized lattice parameters as well as basis
atoms for all structures and models are provided in appendix~\ref{sec:opt_structures}. \par

\begin{figure}[ht!]
	\label{fig:opt_density}
	\centering
	\includegraphics[options]{figures/NiTi_Equilibrium_Density.png}
	\caption{Equilibrium Density for \ce{NiTi} structures with different models.}
\end{figure}

To asses the relative phase stability for a given composition is
to compare the cohesive energy, which is defined as,

\begin{equation}
	\label{eq:cohesive_energy}
	E_{c} = \frac{E_{\text{bulk}} - \sum_{i}^{N} n_i E_{\text{atom}_i}}{N}
\end{equation}

With $E_{\text{bulk}}$ the total energy of the system,
$\sum_i^N n_i E_{\text{atom}_i}$ is the sum over isolated atom
energies for $n_i$ species, and $N$ the number of atoms in bulk
reference. Classical interatomic potentials are offset with respect to
the isolated atom energies, so these terms are zero and neglected,
giving $E_{c} = \frac{E_{\text{bulk}}}{N}$. In fig.~\ref{fig:ecoh} the
bar chart for different phases and interatomic potentials models are
show. We compare these to dervided experimental values for
\textit{B2}\cite{Vandermause2024} and the value from The Materials
Project (MP) entry \texttt{MP-571} which corresponds to the
\textit{B2} phase \cite{MP--571}. In the figure we see that the EAM
and MEAM potentials show significantly lower cohesive energy compared
to the GNN potentials and are in better agreement with the
experimental value. Previous studies using the EAM and MEAM potentials
have been shown to reproduce previous experimental and DFT data
\cite{Haskins2016,Ko2015} such as cohesive energy. The GNNs predict a
more strongly bound and stable \textit{B2} structure but all are
consistent with the entry for The MP and calculation details can be
found in ref.~\cite{MP--571}. This consistency is resonable
considering all pretrained GNN models used in this work have been
developed with MP data. \par

@@

\begin{figure}[!htp]
	\script{PlotPhonons.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_B2_ModelsPhononband structure.png}
		\caption{ Ground-state, no isotropic strain, Phonon bandstructure
			of B2 phase for all models. The Mutter and Zhong EAM potentials
			differ in that Mutter predicts no dynamic instability of the B2
			phase at 0K.  }
		\label{fig:allmodels_b2}
	\end{centering}
\end{figure}

\begin{figure}[!htp]
	\script{PlotPhonons.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_B19_ModelsPhononband structure.png}
		\caption{ Ground-state phonon bandstructure of B19 phase for all
			models.  }
		\label{fig:allmodels_B19P}
	\end{centering}
\end{figure}

\begin{figure}[!htp]
	\script{PlotPhonons.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_B19P_ModelsPhononband structure.png}
		\caption{ Ground-state phonon bandstructure of
			\textit{B19}$^\prime$ phase for all models.  }
		\label{fig:allmodels_B19P}
	\end{centering}
\end{figure}

\begin{figure}[!htp]
	\script{PlotPhonons.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_BCO_ModelsPhononband structure.png}
		\caption{ Ground-state phonon bandstructure of BCO phase for all
			models.  }
		\label{fig:allmodels_B19P}
	\end{centering}
\end{figure}

@@

To better coberate the results in fig.~\ref{fig:modegruneisen}, the
phonon band structure as a result of isoptroic strain applied in
compression and tension are show in fig

\begin{figure}[!htp]
	\script{PlotPhonons.py}
	\begin{centering}
		\includegraphics[width=0.5\linewidth]{figures/NiTi_Mutter_B2_StrainsPhononband structure.png}
		\vspace{1mm}
		\includegraphics[width=0.5\linewidth]{figures/NiTi_Zhong_B2_StrainsPhononband structure.png}
		\vspace{1mm}
		\includegraphics[width=0.5\linewidth]{figures/NiTi_Ko_B2_StrainsPhononband structure.png}
		\caption{ Phonon dispersion of B2 structure for the (top) EAM
			Mutter, (middle) EAM Zhong, and (bottom) 2NN-MEAM Ko potentials
			at different isotropic strains. The optical branches have been
			made more transparent to emphasis the behavior in the TA$_1$ and
			TA$_2$ branches. The color indicates direction from compressive
			to tensile strain. Note that the strain percent shown is with
			respect to isotropic changes (i.e. volume).  }
		\label{fig:mutter_zhong_ko_phonon_b2}
	\end{centering}
\end{figure}

@@

\begin{figure}[!htp]
	\script{PlotPhonons.py}
	\begin{centering}
		\includegraphics[width=0.5\linewidth]{figures/NiTi_M3GNet_B2_StrainsPhononband structure.png}
		\vspace{1mm}
		\includegraphics[width=0.5\linewidth]{figures/NiTi_CHGNet_B2_StrainsPhononband structure.png}
		\vspace{1mm}
		\includegraphics[width=0.5\linewidth]{figures/NiTi_MACE_B2_StrainsPhononband structure.png}
		\caption{ Phonon dispersion of B2 structure for the (top) M3GNet,
			(middle) CHGNet, and (bottom) MACE potentials for different
			isotropic strains. Similar transcparency is used as done in
			\ref{fig:mutter_zhong_phonon_b2}.  }
		\label{fig:gnn_phonon_b2}
	\end{centering}
\end{figure}
